Forecasting the age structure of the scientific workforce in Australia

### 2025-08-21 age_structure_forecasts.qmd
Working on paper

Suppose our workforce is divided into $I$ groups, indexed by $i=1,\dots,I$. In our application, these are scientific disciplines, but in principle they could refer to any subdivision of workers. Let $P_{i,x,t}$ denote the number of equivalent full-time workers in area or discipline $i$ who are aged $x$ at the start of year $t$, where $i=1,\dots,I$ and $x=15,16,\dots$. The starting age of 15 is the minimum age at which individuals are counted as part of the labour force in the Australian Census [@ABSLFSP]. Forecasts are generated for $P_{i,x,T+h}$ across all ages and disciplines, for some forecast horizon $h>0$.

Let $D_{i,x,t}$ denote the number of deaths, and $R_{i,x,t}$ the number of retirements, of people aged $x$ in scientific discipline $i$ in year $t$. This is the number of equivalent full-time workers leaving the workforce during year $t$ who were aged $x$ *at the start* of year $t$.

Two migration processes are required: one representing new Australian workers in the discipline graduating from university courses ($N_{i,x,t}$), and the other representing the net migration process resulting from scientists leaving or arriving in Australia ($G_{i,x,t}$). New workers are treated as a migration process because they can be of any age (unlike births).

The population can be forecasted using a model similar to the stochastic population model of @HB08, excluding the birth process. The demographic growth-balance equation can be written as:
$$
P_{i,x+1,t+1} = P_{i,x,t} - D_{i,x,t} - R_{i,x,t} + N_{i,x,t} + G_{i,x,t}, \quad x = 15,16,\dots
$$ {#eq-dgb}
That is, the population each year is equal to the population from the previous year having aged 1 year, minus the deaths or retirements that occurred during the previous year, plus the new graduates, plus the net migrants (which may be negative). The assumption is that $D_{i,x,t} + R_{i,x,t} \le P_{i,x,t}$ and $N_{i,x,t} = G_{i,x,t} = 0$ above some age threshold (say $x=100$). Once $P_{i,x,t} = 0$ when $x$ is above that threshold, all future populations $P_{i,x+k,t+k} = 0$, $k=1,2,\dots$. That is, when the cohort aged $x$ in year $t$ has all retired or died, they will not be replaced through new graduates or migrants of the same age.

As a first approximation, the components $D$, $R$, $N$ and $G$ can be assumed to behave independently. In reality, there may be a negative correlation between $G$ and $N$ as insufficient graduates would probably lead to employers finding people from overseas, while too many graduates would lead to scientists seeking work elsewhere.

Data on $D$, $N$, and $R$ is required, and the remaining component $G$ will be estimated using @eq-dgb.

To obtain forecasts of $P$, forecasts of $D$, $R$, $N$, and $G$ must then be computed. Each of these can be estimated using a functional data model [@HU07].

@@

## Data Sources

This analysis focuses on the Natural and Physical Sciences as defined in the Australian Standard Classification of Education (ASCED) by the @ABSASCED. For consistency, the report refers to ASCED’s Narrow Fields as "disciplines".

We forecast the number of individuals in each one-year age group who are active in the workforce and hold a bachelor’s degree or higher in the following scientific disciplines: Physics and Astronomy, Mathematical Sciences, Chemical Sciences, Earth Sciences, Biological Sciences, Other Natural and Physical Sciences, and Natural and Physical Sciences not further defined (n.f.d.).

Specifically, the *Census of Population and Housing* (2006, 2011, 2016, and 2021) [@ABSCensus] provides data on the population of individuals with bachelor's degrees or higher who are active in the labour market. This dataset encompasses labour force participation status, age group, educational qualification level (QALLP), discipline (QALFP), and the industries in which individuals work. Labour force participation data is available only for 2016 and 2021 [@ABSLFFP]. QALLP captures the highest level of completed non-school qualification, while QALFP identifies the corresponding field of study [@ABSQALFP].

## Labour Force Participation {#sec-labour}

Labour force participation data is available for 2016 and 2021. To estimate worker numbers for 2006 and 2011, the average participation rates from these later years are applied, assuming overall patterns remain consistent. While participation rates were generally stable between 2016 and 2021, smaller groups experienced more volatility.

Younger and older age groups within some disciplines were particularly affected by this volatility, especially when sample sizes were small. For example, in a group of five 15--19-year-olds, a single individual's employment status change could significantly shift the participation rate. Similarly, fluctuations in smaller cohorts of older scientists impacted participation rates. Averaging the rates from 2016 and 2021 helps mitigate this volatility and provides a more stable estimate for 2006 and 2011. Any discrepancies in participation rates are minimal due to the small cohort sizes, making this a more reliable approach than relying on a single year.

@fig-labour-force displays the labour force participation rates for the broader category of Natural and Physical Sciences for 2016, 2021, and their average, which serves as a proxy for 2006 and 2011. While estimation is conducted separately at both the discipline and category levels, the figure below illustrates this approach using the broader category as an example.

```{r}
#| label: fig-labour-force
#| eval: true
#| fig-cap: Labour force participation rates by one-year age groups for 2016 and 2021, along with the average of these two years, which is used as a proxy for 2006 and 2011.
tar_read(fig1)
```

## Cohort Interpolation {#sec-cohort}

The Census data on scientists active in the labour market, provided by the ABS in one-year age groups, is shown in @fig-census as a line graph to highlight differences across Census years.

As data is only available in Census years (2006, 2011, 2016, 2021), cohort interpolation will be used to estimate intercensal values for the intervening years. These are the years between Census points (e.g., 2007, 2008, 2009, etc.), for which direct population data are unavailable.

Cohort interpolation methods, as described by @STUPP88, approximate these missing values by leveraging available Census data. For this analysis, linear interpolation is applied to estimate the population distribution for the years between Census points, as it provides reasonable approximations given the available data. In this context, linear interpolation produces gradual transitions that reflect realistic population growth and decline.

The completed historical dataset forms the basis for forecasting future workforce trends. The results of this interpolation are shown in @fig-cohort-interpolation.

```{r}
#| label: fig-census
#| eval: true
#| fig-cap: Distribution of scientists active in the labour market by age across  Census years (2006, 2011, 2016, 2021).
tar_read(fig_Pxt_census)
```

```{r}
#| label: fig-cohort-interpolation
#| eval: true
#| fig-cap: Interpolated number of scientists in Australia, 2006--2021, using linear cohort interpolation.
tar_read(fig_Pxt)
```

## Smoothed Disaggregation {#sec-smooth}

@@

## Retirement Distribution by Age {#sec-retirement}

>>>>

Suppose our workforce is divided into $I$ groups, indexed by $i=1,\dots,I$. In our application, these are scientific disciplines, but in principle they could refer to any subdivision of workers. Let $P_{i,x,t}$ denote the number of equivalent full-time workers in group $i$ who are aged $x$ at the start of year $t$, where $x=15,16,\dots$. The starting age of 15 is chosen because it is the minimum age at which individuals are counted as part of the labour force in the Australian Census [@ABSLFSP]. Forecasts are required for $P_{i,x,T+h}$ across all ages and groups, for some forecast horizon $h>0$.

People can leave the workforce of a group through death, retirement, career change, or emigration. They can enter the workforce through graduation, immigration, or career change. Unfortunately, we typically do not have data on many of these processes, so we will combine career change and migration into a remainder term, which we denote as $E_{i,x,t}$. We usually do have data on numbers of deaths ($D_{i,x,t}$), numbers of retirements ($R_{i,x,t}$), and numbers of graduates ($G_{i,x,t}$); the numbers in each case are for people aged $x$ at the start of year $t$.

The population can be forecasted using a model similar to the stochastic population model of @HB08, excluding the birth process:
$$
P_{i,x+1,t+1} = P_{i,x,t} - D_{i,x,t} - R_{i,x,t} + G_{i,x,t} + E_{i,x,t}, \quad x = 15,16,\dots
$$ {#eq-dgb}
That is, the population each year is equal to the population from the previous year having aged 1 year, minus the deaths or retirements that occurred during the previous year, plus the new graduates, plus any other changes due to migration or career change (which may be negative). We assume that $D$, $R$ and $G$ are all non-negative, $D_{i,x,t} + R_{i,x,t} \le P_{i,x,t}$, and $E_{i,x,t} = G_{i,x,t} = 0$ above some age threshold (say $x=100$). Once $P_{i,x,t} = 0$ when $x$ is above that threshold, all future populations $P_{i,x+k,t+k} = 0$, $k=1,2,\dots$. That is, when the cohort aged $x$ in year $t$ has all retired or died, they will not be replaced through new graduates or migrants of the same age.

As a first approximation, the components $D$, $R$, $E$ and $G$ can be assumed to behave independently for each combination of $i$, $x$ and $t$. In reality, there may be a negative correlation between $G$ and $E$ as insufficient graduates would probably lead to employers finding people from overseas, while too many graduates would lead to scientists seeking work elsewhere.

To use @eq-dgb, data on $P$, $D$, $R$, and $G$ is required, and the remainder component $E$ will be calculated using @eq-dgb. It is unlikely that we have available separate death and retirement counts for each group, so we will instead use probabilities of death and retirement across all groups, giving $D_{i,x,t} = P_{i,x,t}q_{x,t}$ and $R_{i,x,t} = P_{i,x,t}r_{x,t}$ where $q_{x,t}$ is the probability of death at each $x$ in year $t$, and $r_{x,t}$ is the probability of retirement at each $x$ in year $t$. Similarly, graduation numbers are rarely available by discipline and age, so we will approximate $G_{i,x,t} = g_xG_t$ where $G_{i,t}$ is the total number of graduates in year $t$ and $g_x$ is the proportion of graduates by age across all disciplines. So our more realistic model becomes
$$
P_{i,x+1,t+1} = P_{i,x,t}(1 - q_{x,t} - r_{x,t}) + g_xG_{i,t} + E_{i,x,t}, \quad x = 15,16,\dots
$$ {#eq-dgb2}
To obtain forecasts of $P$, forecasts of $q$, $r$, $G$, and $E$ must then be computed. Each of these can be estimated using a functional data model [@HU07].

@@

## Data sources

This analysis focuses on the Natural and Physical Sciences as defined in the Australian Standard Classification of Education (ASCED) by the @ABSASCED. We refer to ASCED's Narrow Fields as "disciplines"; these comprise Physics and Astronomy, Mathematical Sciences, Chemical Sciences, Earth Sciences, Biological Sciences, Other Natural and Physical Sciences, and Natural and Physical Sciences not further defined (n.f.d.). We define the population of workers in a discipline as those who are active in the labour market and hold a bachelor's degree or higher in that discipline.

## Labour force participation {#sec-labour}

Data on $P_{i,x,t}$ are sourced from the *Census of Population and Housing* (2006, 2011, 2016, and 2021) [@ABSCensus]. This dataset encompasses labour force participation status, age group, the highest level of completed non-school qualification level (QALLP), the corresponding field of study [QALFP; @ABSQALFP], and the industries in which individuals work.

However, labour force participation data is available only for 2016 and 2021 [@ABSLFFP]. To estimate worker numbers for 2006 and 2011, the average participation rates from 2016 and 2021 were applied, assuming overall patterns remain consistent. While participation rates were generally stable between 2016 and 2021, smaller groups experienced more volatility. Younger and older age groups within some disciplines were particularly affected by this volatility, especially when sample sizes were small. For example, in a group of five 15--19-year-olds, a single individual's employment status change could significantly shift the participation rate. Similarly, fluctuations in smaller cohorts of older scientists impacted participation rates. Averaging the rates from 2016 and 2021 helps mitigate this volatility and provides a more stable estimate for 2006 and 2011. Any discrepancies in participation rates are minimal due to the small cohort sizes, making this a more reliable approach than relying on a single year.

The census data on scientists active in the labour market, provided by the ABS in one-year age groups, is shown in @fig-cohort-interpolation as the thick lines. As data is only available in census years (2006, 2011, 2016, 2021), cohort interpolation is used to estimate values for the intervening years. Cohort interpolation methods [@STUPP88] apply linear interpolation within each age cohort between census years.

```{r}
#| label: fig-cohort-interpolation
#| eval: true
#| fig-cap: Interpolated number of scientists in Australia, 2006--2021, using linear cohort interpolation between census years. Thicker lines are used to denote census years.
tar_read(fig_Pxt)
```

## Retirement distribution {#sec-smooth}

@@
